Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-02-22 ms.tex
editing

Even if SN Ia are rare near the stagnation radius, they may cap the
SMBH accretion rate by periodically blowing gas out of the nucleus of
low mass galaxies.  Between subsequent supernovae, stars release a
gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$
interior to the Ia radius, which is gravitationally bound to the SMBH
by an energy $E_{\rm bind} \sim M_{\rm g}\sigma^{2}$.  Thus, according
to the above definitions, $E_{\rm Ia}/E_{\rm bind} \sim (v_{w}^{\rm
  Ia})^{2}/2\sigma^{2} \gtrsim 1$.  Hence, SN Ia are in principle
capable of dynamically clearing out gas from radii $\sim r_{\rm Ia}
\gtrsim r_{\rm s}$, at least for low mass BHs with $\sigma \ll v^{\rm
  Ia}_{w}$.  Thus, even in cases when heating is sufficiently weak
that the stagnation radius formally exceeds $r_{\rm Ia}$, the
accretion rate is limited to a maximum value,
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{R_{\rm Ia}}{R_{\rm soi}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}]) for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$ (eq.~[$\ref{eq:eddr_analytic}$]).  Note that $\dot{M}_{\rm Ia}$ usually exceeds the maximum accretion rate for a thermally stable flow $\dot{M}_{\rm TI}$ (eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe Ia cannot alone prevent cooling instabilities in the nuclei of low mass galaxies.  Also note that Ia supernovae only cap the accretion rate to this maximum value if the gas feeding the black hole is supplied by local stellar mass loss: the gaseous mass can become much greater if the nucleus is fed externally, e.g. due to a galaxy merger or to cooling flows developing within the galaxy over much longer timescales (e.g.~\citealt{Ciotti&Ostriker07}).

>>>>

Even if SN Ia are rare near the stagnation radius, they may cap the
SMBH accretion rate by periodically blowing gas out of the nucleus of
low mass galaxies.  Between subsequent supernovae, stars release a
gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$
interior to the Ia radius, which is gravitationally bound to the SMBH
by an energy $E_{\rm bind} \sim M_{\rm g}\sigma^{2}$.  Thus, according
to the above definitions, $E_{\rm Ia}/E_{\rm bind} \sim (v_{w}^{\rm
  Ia})^{2}/2\sigma^{2} \gtrsim 1$.  Hence, SN Ia are in principle
capable of dynamically clearing out gas from radii $\sim r_{\rm Ia}
\gtrsim r_{\rm s}$, at least for low mass BHs with $\sigma \ll v^{\rm
  Ia}_{w}$.  Thus, even in cases when heating is sufficiently weak
that the stagnation radius formally exceeds $r_{\rm Ia}$, the
accretion rate is limited to a maximum value,
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{R_{\rm Ia}}{R_{\rm soi}g}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}]) for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$ (eq.~[$\ref{eq:eddr_analytic}$]).  Note that $\dot{M}_{\rm Ia}$ usually exceeds the maximum accretion rate for a thermally stable flow $\dot{M}_{\rm TI}$ (eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe Ia cannot alone prevent cooling instabilities in the nuclei of low mass galaxies.  Also note that Ia supernovae only cap the accretion rate to this maximum value if the gas feeding the black hole is supplied by local stellar mass loss: the gaseous mass can become much greater if the nucleus is fed externally, e.g. due to a galaxy merger or to cooling flows developing within the galaxy over much longer timescales (e.g.~\citealt{Ciotti&Ostriker07}).
